--- 
title: "Building and inferring knowledge bases using biomedical text mining"
author: "Jake Lever"
date: "June 2018"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book
documentclass: ubcthesis
classoption: oneside
#classoption: graybox,envcountchap,sectrefs
#bibliography: [01-intro.bib,02-svd.bib,03-versekindred.bib]
bibliography: thesis.bib
biblio-style: apalike
link-citations: yes
github-repo: jakelever/thesis
description: "This is my PhD thesis on biomedical text mining"
---


```{r bibliographyMerge, eval=T, echo=F}
library(dplyr)
library(kableExtra)
library(RefManageR)
BibOptions(check.entries=F)
bibfiles <- list.files(pattern = "\\.bib$")
bibfiles <- grep("thesis.bib",bibfiles,invert=T,value=T)
bib <- ReadBib(bibfiles[1])
for(bibfile in bibfiles[2:length(bibfiles)])
  bib <- suppressWarnings(suppressMessages(merge(bib,ReadBib(bibfile))))

suppressWarnings(suppressMessages(WriteBib(bib, file='thesis.bib',check=F)))
```

# List of Abbreviations

```{r abbreviations, eval=T, echo=F}
abbreviations <- read.table('data/abbreviations.tsv',sep='\t',comment='')

knitr::kable(abbreviations,
  booktabs = TRUE, escape=FALSE,
  longtable=TRUE,
  col.names = c('Short','Long')
  )
```


# Acknowledgements

I would like to thank my Ph.D. supervisor, Dr. Steven J. M. Jones, who provided a wonderful environment to explore text mining and cancer genomics. He encouraged me to pursue countless opportunities to present my work and build collaborations. Through this, I learned a great deal about the practicalities of team science that I hope will benefit my career for years to come. I am very thankful that I chose to do my graduate work with him. Many thanks go to Sharon Ruschkowski and Louise Clarke for making my time in graduate school go so smoothly. I would also like to thank my supervisory committee, Drs. Sohrab Shah, Art Cherkasov, and Inanc Birol, for their support and guidance during this research.

I was very lucky to have the opportunity to conduct research alongside the lead researchers of the Personalized Oncogenomics (POG) program. Sharing an office with Drs. Yaoqing Shen, Erin Pleasance, Martin Jones and Laura Williamson was one of the key factors for my enjoyment of my entire Ph.D. Their friendship, anecdotes, and discussions created a wonderful atmosphere. All the other members of the Jones lab, including Eric Zhao, Jasleen Grewal, Luka Culibrk, Celia Siu and Santina Lin, have been good friends whom I hope to stay connected with as our careers progress. They all contributed invaluable work to this research. I am thankful for all the staff at the Genome Sciences Centre and involved in the Personalized Oncogenomics (POG) project for the incredible work that continues to be done. The other members of my Bioinformatics program cohort, including Shaun Jackman and Sarah Perez, have been great friends who helped me settle easily into the program and life in Canada.

I offer Martin Krzywinski great thanks for our conversations and the fantastic opportunities to contribute to the Points of Significance column and several visualization projects. These were wonderful learning experiences in collaboration and communication. His leadership of the coffee club at the Genome Sciences Centre was also an essential component of my research success. I am also thankful for the many conversations with Dr. Morgan Bye and Amir Zadeh.

I would also like to thank Drs. Obi Griffith, Malachi Griffith, Kilannin Krysiak, Arpad Danos and other members of the Griffith lab at Washington University at St. Louis for all their support and also their hard work that provided data for this work. I would like to thank Dr. Ben Busby at the National Center of Biotechnology Information (NCBI) for his encouragement in my research and for welcoming me to Bethesda for a short research placement during my Ph.D. as a visiting bioinformatician.

Many thanks go to the various funding agencies that have supported this research. This includes a Vanier Canada Graduate Scholarship, a UBC Four Year Fellowship, a scholarship from the MSFHR/CIHR Bioinformatics training program and funding from the OpenMinTeD Horizons 2020 project. The BC Cancer Foundation supports the POG project which has further enabled this research. This work would not have been possible without the incredible online community of developers and researchers who were willing to answer questions, particularly at StackOverflow and RStudio.

Lastly, I would like to thank my family and my partner Chantal for the unending support that I have received.

# Dedication

To my mother Ann, my late father David, my sister Sarah and my partner Chantal.

\mainmatter